---
title: "Empowering Women's Wellness: Unveiling Smart Device Insights for Bellabeat's Growth and Marketing Strategy"
author: "Prepared by: Mr. Jefferson Mwatati"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: cayman
  html_document:
    toc: true
---


```{r}
#All required packages

#install.packages("tidyverse")
#install.packages("dplyr") #helps in data manipulation
#install.packages("here")  #makes referencing files easy
#install.packages("skimr") #good for summarizing data easily
#install.packages("janitor") #it has functions for cleaning data
#install.packages("prettydoc")

```

```{r}
#Load all installed packages

library(tidyverse)
library(dplyr)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(knitr)
library(prettydoc)

```



```{r}
# import data
dailyactivity_df <- read_csv("dailyActivity_merged.csv")
```
View first 6 records.

```{r}
#view data
head(dailyactivity_df)
```


```{r}
#import data
sleepday_df <- read_csv("sleepDay_merged.csv")
```




```{r}
#view data
head(sleepday_df)
```




```{r}
#import data
weightloginfo_df <- read_csv("weightLogInfo_merged.csv")
```



```{r}
#view data
head(weightloginfo_df)
``` 

```{r}
#View column names and datatypes 
glimpse(dailyactivity_df)
```

```{r}
# create new column
newdailyactivity_df <- dailyactivity_df %>%
  mutate(newActivityDate = mdy(ActivityDate))
```


```{r}
#select and arrange columns
cleandailyactivity_df <- select(newdailyactivity_df,Id, newActivityDate,TotalSteps,TotalDistance,Calories)
```

```{r}
#view data set
head(cleandailyactivity_df)
```


```{r}
# Calculate statistics
activity_totalrecords <- nrow(cleandailyactivity_df)
activity_totaldays <- length(unique(cleandailyactivity_df$newActivityDate))
activity_totalparticipants <- length(unique(cleandailyactivity_df$Id))
average_daily_steps <- mean(cleandailyactivity_df$TotalSteps)
average_daily_distance <- mean(cleandailyactivity_df$TotalDistance)
average_daily_calories_burned <- mean(cleandailyactivity_df$Calories)

# Print the calculated statistics
cat("Total Activity Records:", activity_totalrecords, "\n")
cat("Total Activity Days:", activity_totaldays, "\n")
cat("Total women participants:", activity_totalparticipants, "\n")
cat("Average of daily steps made:", average_daily_steps, "\n")
cat("Average of daily distance covered:", average_daily_distance, "\n")
cat("Average of daily calories burned:", average_daily_calories_burned, "\n")
```



```{r}
# Determine activity first day and last day
activitydaterange_df <- cleandailyactivity_df %>%
  summarize(
    activity_firstday = min(newActivityDate),
    activity_lastday = max(newActivityDate)
  )

# Display results
head(activitydaterange_df)
```


```{r}
# Calculate the trend in total steps over time
total_steps_by_date <- cleandailyactivity_df %>%
  group_by(newActivityDate) %>%
  summarise(total_steps = sum(TotalSteps))

total_steps_by_date
```


```{r}
# Calculate the trend in total distance over time
total_distance_by_date <- cleandailyactivity_df %>%
  group_by(newActivityDate) %>%
  summarise(total_distance = sum(TotalDistance))
total_distance_by_date
```


```{r}
# Calculate the trend in calories burned over time
calories_burned_by_date <- cleandailyactivity_df %>%
  group_by(newActivityDate) %>%
  summarise(calories_burned = sum(Calories))

calories_burned_by_date
```


```{r}
#Filtering a dataset using the pipe operator to extract specific records
fitnesswoman_001 <- cleandailyactivity_df %>%
  filter(Id=="4020332650") 
```




```{r}
#Filtering a dataset using the pipe operator to extract specific records
fitnesswoman_002 <- cleandailyactivity_df %>%
  filter(Id=="4388161847") 
```






```{r}
#Filtering a dataset using the pipe operator to extract specific records
fitnesswoman_003 <- cleandailyactivity_df %>%
  filter(Id=="2347167796") 
```




```{r}
#Filtering a dataset using the pipe operator to extract specific records
fitnesswoman_005 <- cleandailyactivity_df %>%
  filter(Id %in% c(1503960366, 1624580081, 1844505072, 1927972279, 2022484408, 2026352035, 2320127002, 2347167796, 2873212765, 3372868164, 3977333714, 4020332650))

```






```{r}
glimpse(sleepday_df)
```



```{r}
# Convert the "SleepDay" column to a proper date-time format
sleepday_df$SleepDay <- as.POSIXct(sleepday_df$SleepDay, format = "%m/%d/%Y %I:%M:%S %p")

# Extract the date and time from the converted date-time column
sleepday_df$Date <- as.Date(sleepday_df$SleepDay)
sleepday_df$Time <- format(sleepday_df$SleepDay, "%H:%M:%S")

```


```{r}
#select and arrange columns
cleansleepday_df <- select(sleepday_df,Id,Date,Time,TotalSleepRecords,TotalMinutesAsleep,TotalTimeInBed)
```


```{r}
#view data
head(cleansleepday_df)
```


```{r}
# Calculate statistics
sleep_totalrecords <- nrow(cleansleepday_df)
sleep_totaldays <- length(unique(cleansleepday_df$Date))
sleep_totalwomen_participants <- length(unique(cleansleepday_df$Id))
minimum_totalminutesasleep <- min(cleansleepday_df$TotalMinutesAsleep)
maximum_totalminutesasleep <- max(cleansleepday_df$TotalMinutesAsleep)
average_totalminutesasleep <- mean(cleansleepday_df$TotalMinutesAsleep)
minimum_TotalTimeInBed <- min(cleansleepday_df$TotalTimeInBed)
maximum_TotalTimeInBed <- max(cleansleepday_df$TotalTimeInBed)
average_TotalTimeInBed <- mean(cleansleepday_df$TotalTimeInBed)

# Print the calculated statistics
cat("Total sleep records:", sleep_totalrecords, "\n")
cat("Total sleep days recorded:", sleep_totaldays, "\n")
cat("Total women participants in sleep records:", sleep_totalwomen_participants, "\n")
cat("Minimum minutes asleep :", minimum_totalminutesasleep, "\n")
cat("Maximum minutes asleep:", maximum_totalminutesasleep, "\n")
cat("Average minutes asleep:", average_totalminutesasleep, "\n")
cat("Minimum time in bed:", minimum_TotalTimeInBed, "\n")
cat("Maximum time in bed:", maximum_TotalTimeInBed, "\n")
cat("Average time in bed:", average_TotalTimeInBed, "\n")
```



```{r}
# Calculate the correlation coefficient between total minutes asleep and total time in bed
cor(cleansleepday_df$TotalMinutesAsleep, cleansleepday_df$TotalTimeInBed)
```



```{r}
# Check the correlation between numerical variables
cor(cleansleepday_df[, c("TotalSleepRecords", "TotalMinutesAsleep", "TotalTimeInBed")])

```


```{r}
# Calculate the trend in total minutes asleep over time
TotalMinutesAsleep_by_date <- cleansleepday_df %>%
  group_by(Date) %>%
  summarise(total_minutes_asleep = sum(TotalMinutesAsleep))

TotalMinutesAsleep_by_date
```




```{r}
# Calculate the trend in total time in bed over time
TotalTimeInBed_by_date <- cleansleepday_df %>%
  group_by(Date) %>%
  summarise(total_time_in_bed = sum(TotalTimeInBed))

TotalTimeInBed_by_date
```


```{r}
#Filter specific records
fitnesswomenTTIB <- cleansleepday_df %>%
  filter(Id %in% c(8378563200))

```




```{r}
glimpse(weightloginfo_df)
```




```{r}
#Change the date column datatype from chr to datetime
weightloginfo_df$Date <- as.POSIXct(weightloginfo_df$Date, format = "%m/%d/%Y %I:%M:%S %p")

#Create a column for date only
weightloginfo_df$DateOnly <- as.Date(weightloginfo_df$Date)

#Create a column for Time only
weightloginfo_df$TimeOnly <- format(weightloginfo_df$Date, format = "%H:%M:%S")


```



```{r}
#select and arrange columns
cleanweightloginfo_df <- select(weightloginfo_df,Id,DateOnly,TimeOnly,WeightKg,WeightPounds,BMI,IsManualReport,LogId)
```




```{r}
#View data
head(cleanweightloginfo_df)
```




```{r}
# Calculate statistics
weightinfo_totalrecords <- nrow(cleanweightloginfo_df)
weightinfo_totaldays <- length(unique(cleanweightloginfo_df$DateOnly))
weightinfo_totalparticipants <- length(unique(cleanweightloginfo_df$Id))
mean_weight <- mean(cleanweightloginfo_df$WeightKg)
min_weight <- min(cleanweightloginfo_df$WeightKg)
max_weight <- max(cleanweightloginfo_df$WeightKg)


# Print the calculated statistics
cat("Total Weightinfo Records:", weightinfo_totalrecords, "\n")
cat("Total Weightinfo collection Days:", weightinfo_totaldays, "\n")
cat("Total women participants:", weightinfo_totalparticipants, "\n")
cat("Average Weight:", mean_weight, "\n")
cat("Minimum Weight:", min_weight, "\n")
cat("Maximum Weight:", max_weight, "\n")

```



```{r}
correlation <- cor(cleanweightloginfo_df$WeightKg, cleanweightloginfo_df$BMI)
cat("Correlation between Weight and BMI:", correlation, "\n")

```




```{r}
# Group the data and calculate the counts
grouped_IsManualReport <- cleanweightloginfo_df %>%
  group_by(IsManualReport) %>%
  summarize(no_of_reports = n())

grouped_IsManualReport
```


```{r}
# Create a new column with the BMI categories
BMI_category_Count <- cleanweightloginfo_df %>%
  mutate(BMI_Category = case_when(
    BMI < 18.5 ~ "Underweight",
    BMI >= 18.5 & BMI < 25 ~ "Normal weight",
    BMI >= 25 & BMI < 30 ~ "Overweight",
    BMI >= 30 & BMI < 35 ~ "Obesity",
    BMI >= 35 & BMI < 40 ~ "Class I Obesity",
    BMI >= 40 & BMI < 45 ~ "Class II Obesity",
    BMI >= 45 ~ "Class III Obesity"
  ))

# Count the occurrences of each BMI category
counted_categories <- BMI_category_Count %>%
  group_by(BMI_Category) %>%
  summarise(Count = n())

counted_categories

```




```{r}
# Plot the trend in total steps over time
activity_firstday <- min(cleandailyactivity_df$newActivityDate)
activity_lastday <- max(cleandailyactivity_df$newActivityDate)

ggplot(total_steps_by_date, aes(newActivityDate, total_steps)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  labs(title = "Tracking Progress: Total Steps Trend Over Time",
       subtitle=paste0("Analyzing Activity Data from: ", activity_firstday, " to ", activity_lastday),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Activity Date",
       y="Total Steps")

```



```{r}
# Plot the trend in total distance over time
ggplot(total_distance_by_date, aes(newActivityDate, total_distance)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  labs(title = "Total Distance Trend Analysis",
       subtitle=paste0("Exploring Changes in Total Distance Covered from: ", activity_firstday, " to ", activity_lastday),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Activity Date",
       y="Total Distance")

```



```{r}
# Plot the trend in calories burned over time
ggplot(calories_burned_by_date, aes(newActivityDate, calories_burned)) +
  geom_line(color = "purple", linewidth = 1.2) +
  labs(title = "Tracking Energy Expenditure: Trend in Calories Burned Over Time",
       subtitle=paste0("Analyzing Activity Data from: ", activity_firstday, " to ", activity_lastday),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Activity Date",
       y="Calories Burned")
```



```{r}
ggplot(fitnesswoman_001, aes(newActivityDate, TotalSteps)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  labs(title = "Tracking Total Steps: Unveiling the 31-Day Journey",
       subtitle=paste0("A Comprehensive Trend Analysis of Participant's Activity"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Activity Date",
       y="Total Steps")
```





```{r}
ggplot(fitnesswoman_002, aes(newActivityDate, TotalSteps)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  labs(title = "Exploring Activity Fluctuations: A 31-Day Step Journey",
       subtitle=paste0("Analyzing Variations in Total Steps for a Month"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Activity Date",
       y="Total Steps")
```



```{r}
ggplot(fitnesswoman_003, aes(newActivityDate, TotalSteps)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  labs(title = "Monthly Total Steps Analysis: Unveiling Activity Trends",
       subtitle=paste0("Exploring 31 Days of Physical Activity Patterns"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Activity Date",
       y="Total Steps")
```   



```{r}
ggplot(fitnesswoman_005, aes(newActivityDate, TotalSteps)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  facet_wrap(~Id) +
  labs(title = "Individual Total Steps Analysis: A Multifaceted View",
       subtitle=paste0("Exploring 31 Days of Activity Trends for Different Women"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Activity Date",
       y="Total Steps")
```



```{r}
# Plot the trend in total minutes asleep over time

ggplot(TotalMinutesAsleep_by_date, aes(Date, total_minutes_asleep)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  labs(title = "Exploring Sleep Patterns: Total Minutes Asleep Over Time",
       subtitle=paste0("Analyzing Sleep Trends Across Dates for 24 Women"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Date",
       y="Total Minutes Asleep")

```



```{r}
# Plot the trend in total time in bed over time

ggplot(TotalTimeInBed_by_date, aes(Date, total_time_in_bed)) +
  geom_line(color = "darkred", linewidth = 1.2) +
  labs(title = "Tracking Bedtime Trends: Total Time in Bed Over a Month",
       subtitle=paste0("Analyzing Bedtime Duration Patterns Across Dates for 24 Women"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Date",
       y="Total Time in Bed")

```



```{r}
ggplot(fitnesswomenTTIB, aes(Date,TotalTimeInBed)) +
  geom_line(color = "maroon", linewidth = 1.2) +
  labs(title = "Exploring Bedtime Patterns: 31-Day Total Time in Bed Analysis",
       subtitle=paste0("Tracing a Participant's Bedtime Trends Over Time"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"),
       x="Date",
       y="Total Time in Bed")+
  theme_light()

```




```{r}
# Create a histogram to visualize the distribution of TotalMinutesAsleep

ggplot(cleansleepday_df, aes(x = TotalMinutesAsleep)) +
  geom_histogram(binwidth = 30, col = "white", fill = "darkred") +
  labs(title = "Understanding Sleep Duration Distribution",subtitle=paste0("Exploring the Spread of Total Minutes Asleep in a Dataset"),x = "Minutes Asleep", y = "Frequency", caption=paste0("Prepared by Mr. Jefferson Mwatati")) +
  theme_bw()

```



```{r}
# Plot the distribution of total time in bed
ggplot(cleansleepday_df, aes(x = TotalTimeInBed)) +
  geom_histogram(binwidth = 10, col = "white", fill = "darkblue") +
  labs(title = "Exploring Bedtime Duration Distribution",subtitle=paste0("Analyzing the Spread of Total Time in Bed within the Dataset"), x = "Total Time in Bed (minutes)", y = "Count", caption=paste0("Prepared by Mr. Jefferson Mwatati")) +
  theme_light()
```


```{r}
# Create a scatter plot with regression line
ggplot(cleansleepday_df, aes(x = TotalMinutesAsleep, y = TotalTimeInBed)) +
  geom_point(color = "brown") +
  geom_smooth(method = "lm", se = FALSE,linewidth = 1.4, color = "green") +
  labs(x = "Total Minutes Asleep", y = "Total Time in Bed (minutes)", title = "Exploring the Relationship between Sleep Duration and Bedtime",subtitle=paste0("A strong positive correlation exists between total minutes asleep and total time in bed"), caption=paste0("Prepared by Mr. Jefferson Mwatati"))+
   theme_light()
  

```


```{r}
ggplot(cleanweightloginfo_df, aes(x = WeightKg)) +
  geom_histogram(binwidth = 9, fill = "blue", color = "black") +
  labs(x = "Weight (Kg)", 
       y = "Frequency", 
       title = "Distribution of Weight in Kilograms",
       subtitle=paste0("Frequency Distribution of Recorded Weights"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati"))

```


```{r}
# Create a pie chart with percentages
ggplot(grouped_IsManualReport, aes(x = "", y = no_of_reports, fill = factor(IsManualReport))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(fill = "IsManualReport",
       title = "Manual and Automatic Reports",
       subtitle=paste0("Proportion of Reports Generated Manually and Automatically"),
       caption=paste0("Prepared by Mr. Jefferson Mwatati")) +
  scale_fill_manual(values = c("TRUE" = "purple", "FALSE" = "orange")) +
  geom_text(aes(label = scales::percent(no_of_reports / sum(no_of_reports))), 
            position = position_stack(vjust = 0.5)) +
  theme_void()


```
 

```{r}
# Create a pie chart
ggplot(counted_categories, aes(x = "", y = Count, fill = BMI_Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(fill = "BMI Category",
       title = "BMI Category Distribution",
       subtitle=paste0("Proportion of Different BMI Categories Among Individuals"),
       caption = "Prepared by Mr. Jefferson Mwatati") +
  scale_fill_brewer(palette = "Set1") +  
  geom_text(aes(label = scales::percent(Count / sum(Count))), 
            position = position_stack(vjust = 0.5)) +
  theme_void()
```






## **A summary of analysis**

The analysis of the smart device data has revealed some interesting trends that could be applied to Bellabeat customers and inform their marketing strategy.Here are the key findings based on the analysis:

<div style="color: blue;">**There is a decline in activity levels over time.**</div> This is evident in the total steps taken, total distance covered, and calories burned data. This trend could be due to a number of factors, such as busy schedules, changes in weather, or simply boredom with the same old workout routine. Bellabeat could target their marketing efforts to encourage users to stay active, even when they're feeling unmotivated. They could offer new workout challenges, provide access to workout classes, or simply send motivational messages to their users.

<div style="color: blue;">**There is a lot of variability in sleep patterns.**</div> This is evident in the total minutes asleep and total time in bed data. Some users get a consistent 8 hours of sleep per night, while others have much more variable sleep patterns. Bellabeat could target their marketing efforts to help users improve their sleep quality. They could offer tips on sleep hygiene, provide access to sleep tracking apps, or even sell sleep products.

<div style="color: blue;">**There is a strong correlation between weight and BMI.**</div> This means that as weight increases, BMI also increases. This is to be expected, as BMI is a measure of body fat, and weight is a major component of body fat. Bellabeat could use this information to target their marketing efforts to users who are overweight or obese. They could offer weight loss programs, provide access to healthy recipes, or even sell fitness trackers.

<div style="color: blue;">**A large proportion of reports are manual.**</div> This suggests that there is a need for Bellabeat to improve their automated data collection process. They could invest in new software or hire more staff to help with data entry. This would free up their users to focus on their health and wellness goals, rather than spending time entering data manually.

<div style="color: blue;">**The majority of participants are classified as "Normal weight" or "Overweight."**</div> This suggests that Bellabeat has a good mix of users, from those who are looking to maintain a healthy weight to those who are looking to lose weight. Bellabeat could target their marketing efforts to both groups of users, offering different products and services to meet their individual needs.


#### **Influence on Marketing Strategy**

* **Activity-focused Campaigns:** Bellabeat can run campaigns promoting the benefits of consistent physical activity, highlighting the connection between activity levels and overall health. This can attract users who are looking for holistic health solutions.

* **Sleep Improvement Initiatives:** Marketing efforts can emphasize the importance of quality sleep and how Bellabeat's products can help users achieve better sleep patterns. Testimonials and success stories can reinforce these messages.

* **Personalized Weight Management:** Bellabeat can position itself as a comprehensive health companion by showcasing how its products help users manage their weight and BMI effectively. Educational content on weight management can establish the company as an authority in the field.

		

## **Recommendations Based on Analysis**

Here are the recommendations based on the analysis:

* **Create a challenge to encourage users to increase their activity levels.** This could be a step challenge, a distance challenge, or a calorie burn challenge. Bellabeat could offer prizes to the winners of the challenge to motivate users to participate.
* **Offer a sleep tracking app or program.** This would help users track their sleep patterns and identify areas where they can improve their sleep quality. Bellabeat could also offer tips on sleep hygiene and provide access to relaxing music or sounds to help users fall asleep.
* **Create a weight loss program.** This could include meal plans, workout routines, and support from a coach. Bellabeat could also offer discounts on fitness trackers or other products that can help users lose weight.
* **Improve their automated data collection process.** This would free up their users to focus on their health and wellness goals, rather than spending time entering data manually.
* **Target their marketing efforts to specific groups of users.** This could include users who are looking to maintain a healthy weight, users who are looking to lose weight, or users who are interested in improving their sleep quality.
* **Offer products and services that meet the individual needs of their users.** This could include fitness trackers, workout clothes, meal plans, or weight loss programs.
* **Add more data points to monitor users behavior.** This could include age,diet of food eaten every day, residence etc. This kind of data points will add to the analysis depth and understanding of the users.


#### **Recommendations for New Fitness Clients:**

* **Activity:** New fitness clients should aim for a daily step count and distance similar to the analyzed averages. Gradual increases in activity can lead to improved health outcomes.

* **Sleep:** Clients should focus on maintaining consistent sleep schedules and durations. Prioritizing sleep quality through relaxation techniques and sleep hygiene practices is essential.

* **Weight Management:** Clients should be educated about the correlation between weight and BMI. A balanced diet, regular exercise, and consultation with healthcare professionals can aid in achieving and maintaining a healthy weight.


In conclusion, the analysis of smart device data provides valuable insights for Bellabeat's growth strategy. By tailoring their products to address the identified trends and user needs, Bellabeat can enhance customer engagement and solidify its position as a leader in women's health technology. Additionally, implementing the recommended marketing strategies can help the company effectively communicate its value proposition to its target audience.

################################################################################
                                    **End of Report**




















